Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-05-10 ppdp19/ppdp19.tex
[ ppdp19 ] quick reading of Section 4
[ ppdp19 ] direct links to Agda HTML for sections 2 and 4

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen,fleqn]{acmart}

@@

Our design choices were guided by an Agda formalization of sections 2
(complete) and 4 (partial), available at \url{https://github.com/andreasabel/ipl}.
Agda was particularly helpful to correctly handle the renamings
abundantly present when working with presheaves.

@@

Due to the eager splitting of positive hypotheses, reflection at type $P$
now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
produce a single positive semantic value; instead, one such value is needed
for every branch.  We implement
\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
as a higher-order function expecting a continuation $k$ which is
invoked for each generated branch with the semantic value of type $P$
constructed for this branch.
\[
\renewcommand{\left}{}
\renewcommand{\right}{}
\begin{array}{l@{~}lcl}
  \reflect {\patom} & k & = &
    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
  \reflect {\Thunk N} & k & = &
    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
  \reflect {\tyzero} & k & = & \nbranch \\
  \reflect {P_1 + P_2} & k & = & \bbranch
    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
\\ &&& \qquad \quad \ \,
    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
    \\
  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
  \reflect{P_1 \times P_2} & k & = & \bsplit
    \,\left(
      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
\\ &&& \qquad \ \
      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
\\ &&& \qquad \ \
        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
        \right)
      \right)
    \right)
\end{array}
\]
Reflecting at a positive atomic type $\patom$ is the regular ending of a
reflection pass: we call continuation $k$ with a fresh variable
$\pvar\,\tzero$ of type $\patom$, making space for the variable
using $\twk^{\patom}$.
In case we end at type $\Thunk N$, we add a new variable
$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
$\eta$-expansion via $\up N$.  Two more endings are possible:
At type
$\tyzero$, we have reached an absurd case, meaning that no
continuation is necessary since we can conclude with
\emph{ex falsum quod libet}.
At type $\tyone$, there is no need to add a new variable, as values
of type $\tyone$ contain no information.  We simply pass the unit
value $\ttempty$ to $k$ in this case.
Reflecting at $P_1 + P_2$ generates two branches, which may result in
several uses of the continuation $k$.
In the first branch, we recursively reflect at $P_1$.  Its
continuation will receive a semantic value in $\den{P_1}$ which we
inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
second branch proceeds analogously.
Finally reflecting at $P_1 \times P_2$ means we first have to analyze
$P_1$, and in each of the generated branches we continue to analyze
$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
$\reflect{P_1}$.  Each reflection phase gives us a semantic value
$a_i$ of type $P_i$, which we combine to a tuple before passing it to
$k$.  Note also that the context extension $\tau_1$ created in the
first phase needs to be composed with the context extension $\tau_2$
of the second phase to transport $k$ into the final context.  Further,
the value $a_1$ was constructed relative to the target of
$\tau_1$ and still needs
to be transported with $\tau_2$ before paired up with $a_2$.

The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
reflection and reification at negative types.
% and identity environment creation.
\[
\begin{array}{l@{~}lcl}
  \downG {P \To N} & f & = & \tabs
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
          \down[\Delta] P (f \; \tau \; a)
        \right)
    \right)
  \\
  \upG {\Comp P} & u & = & \tbind \; u
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \; \tau \; a .\
          \treturn\; a
        \right)
    \right)
  % \\
  % \idenv{\cext \Gamma P} & = &
\end{array}
\]
% To create an identity environment using $\treflect$, let us first
% extend $\Add\_$ to contexts, following the intuition that
%Due to the strong limitation on context entries,
Due to the absence of composite positive types in contexts,
the identity environment $\idenv \Gamma$ can be built
straightforwardly using negative reflection.
\[
\begin{array}{lcl}
  \idenv{\cempty} & = & \ttempty \\
  \idenv{\cext \Gamma {\patom}} & = &
    (
      \tren\;\twk^{\patom}\;\idenv\Gamma
      , \
      \pvar\,\tzero
    ) \\
  \idenv{\cext \Gamma N} & = &
    (
      \tren\;\twk^N\;\idenv\Gamma
      , \
      \upG N (\nvar\,\tzero)
    ) \\
\end{array}
% \begin{array}{lcl}
%   \idenv{\cempty} & = & \ttempty \\
%   \idenv{\cext \Gamma {\patom}} & = &
%     \left(
%       \tren\;\twk^{\patom}\;\idenv\Gamma
%       , \
%       \pvar\,\tzero
%     \right) \\
%   \idenv{\cext \Gamma N} & = &
%     \left(
%       \tren\;\twk^N\;\idenv\Gamma
%       , \
%       \upG N (\nvar\,\tzero)
%     \right) \\
% \end{array}
\]

@@

We have implemented NbE for CBPV and polarized lambda calculus
formulated with intrinsically well-typed syntax and presheaf
semantics.  As a side result, we have proven semantically
that the normal forms of both systems are logically complete, \ie,
each derivable judgement $\Gamma \der N$ has a normal derivation.
It remains to show that NbE for these calculi is also computationally
sound and complete, i.e., the computational behavior of term and
normal form should agree, and normalization should decide a suitable
equational theory on terms.

>>>>

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen,fleqn,review]{acmart} % review for draft
%% DRAFT only:
\settopmatter{printacmref=false} % Removes citation information below abstract
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
\sloppy

@@

Our design choices were guided by an Agda formalization of sections 2
(complete, see \url{https://andreasabel.github.io/ipl/html/NfModelMonad.html})
and 4 (partial, see \url{https://andreasabel.github.io/ipl/html-focusing/Polarized.html}).
Agda was particularly helpful to correctly handle the renamings
abundantly present when working with presheaves.

@@

Due to the eager splitting of positive hypotheses, reflection at type $P$
now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
produce a single positive semantic value; instead, one such value is needed
for every branch.  We implement
\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
as a higher-order function expecting a continuation $k$ which is
invoked for each generated branch with the semantic value of type $P$
constructed for this branch.
\[
\renewcommand{\left}{}
\renewcommand{\right}{}
\begin{array}{l@{~}lcl}
  \reflect {\patom} & k & = &
    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
  \reflect {\Thunk N} & k & = &
    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
  \reflect {\tyzero} & k & = & \nbranch \\
  \reflect {P_1 + P_2} & k & = & \bbranch
    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
\\ &&& \qquad \quad \ \,
    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
    \\
  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
  \reflect{P_1 \times P_2} & k & = & \bsplit
    \,\left(
      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
\\ &&& \qquad \ \
      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
\\ &&& \qquad \ \
        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
        \right)
      \right)
    \right)
\end{array}
\]
Reflecting at a positive atomic type $\patom$ is the regular ending of a
reflection pass: we call continuation $k$ with a fresh variable
$\pvar\,\tzero$ of type $\patom$, making space for the variable
using $\twk^{\patom}$.
In case we end at type $\Thunk N$, we add a new variable
$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
$\eta$-expansion via $\up N$.  Two more endings are possible:
At type
$\tyzero$, we have reached an absurd case, meaning that no
continuation is necessary since we can conclude with
\emph{ex falso quodlibet}.
At type $\tyone$, there is no need to add a new variable, as values
of type $\tyone$ contain no information.  We simply pass the unit
value $\ttempty$ to $k$ in this case.
Reflecting at $P_1 + P_2$ generates two branches, which may result in
several uses of the continuation $k$.
In the first branch, we recursively reflect at $P_1$.  Its
continuation will receive a semantic value in $\den{P_1}$ which we
inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
second branch proceeds analogously.
Finally reflecting at $P_1 \times P_2$ means we first have to analyze
$P_1$, and in each of the generated branches we continue to analyze
$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
$\reflect{P_1}$.  Each reflection phase gives us a semantic value
$a_i$ of type $P_i$, which we combine to a tuple before passing it to
$k$.  Note also that the context extension $\tau_1$ created in the
first phase needs to be composed with the context extension $\tau_2$
of the second phase to transport $k$ into the final context.  Further,
the value $a_1$ was constructed relative to the target of
$\tau_1$ and still needs
to be transported with $\tau_2$ before paired up with $a_2$.

The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
reflection and reification at negative types.
% and identity environment creation.
\[
\begin{array}{l@{~}l@{~~}c@{~~}l}
  \downG {P \To N} & f & = & \tabs
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \, _\Delta \, (\tau : \Gamma \sublist \Delta) \; a .\
          \down[\Delta] P (f \; \tau \; a)
        \right)
    \right)
  \\
  \upG {\Comp P} & u & = & \tbind \; u
    \left(
      \reflect[\Gamma] P
        \left(
          \lambda \; \tau \; a .\
          \treturn\; a
        \right)
    \right)
  % \\
  % \idenv{\cext \Gamma P} & = &
\end{array}
\]
% To create an identity environment using $\treflect$, let us first
% extend $\Add\_$ to contexts, following the intuition that
%Due to the strong limitation on context entries,
Due to the absence of composite positive types in contexts,
the identity environment $\idenv \Gamma$ can be built
straightforwardly using negative reflection.
\[
\begin{array}{lcl}
  \idenv{\cempty} & = & \ttempty \\
  \idenv{\cext \Gamma {\patom}} & = &
    (
      \tren\;\twk^{\patom}\;\idenv\Gamma
      , \
      \pvar\,\tzero
    ) \\
  \idenv{\cext \Gamma N} & = &
    (
      \tren\;\twk^N\;\idenv\Gamma
      , \
      \upG N (\nvar\,\tzero)
    ) \\
\end{array}
% \begin{array}{lcl}
%   \idenv{\cempty} & = & \ttempty \\
%   \idenv{\cext \Gamma {\patom}} & = &
%     \left(
%       \tren\;\twk^{\patom}\;\idenv\Gamma
%       , \
%       \pvar\,\tzero
%     \right) \\
%   \idenv{\cext \Gamma N} & = &
%     \left(
%       \tren\;\twk^N\;\idenv\Gamma
%       , \
%       \upG N (\nvar\,\tzero)
%     \right) \\
% \end{array}
\]

@@

We have defined NbE for CBPV and polarized lambda calculus,
formulated with intrinsically well-typed syntax and presheaf
semantics.  As a side result, we have proven semantically
that the normal forms of both systems are logically complete, \ie,
each derivable judgement $\Gamma \der N$ has a normal derivation.
It remains to show that NbE for these calculi is also computationally
sound and complete, i.e., the computational behavior of term and
normal form should agree, and normalization should decide a suitable
equational theory on terms.
